---
title: "An Analysis of the Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Clinton Olson"
date: "May 8, 2016"
output: pdf_document
---

```{r setup, echo = FALSE}
library(knitr)
library(ggplot2)
library(gridExtra)
data("ToothGrowth")
```

#Overview
This analysis examines the effect of vitamin C dosage and delievery method on the length of odontoblast
cells in 60 guinea pigs from the ToothGrowth dataset supplied by R.  Here we examined the effect of 
increasing dosages of vitamin C on cell length and found that larger dosages (up to 2.0 mg/day) 
correspond with significantly greater mean cell length.  We also examined the effect of delivery method
(orange juice vs. ascorbic acid) on mean cell length and found a significantly greater mean length
for guinea pigs receiving vitamin C via orange juice.

#Data Summary
The data for this analysis comes from the ToothGrowth dataset preloaded into R.  There are 60 
observations in this data where each observation includes the vitamin C dosage, 
delivery method (one of orange juice (OJ) or ascorbic acid (VC)), and the length of the odontoblast cells
responsible for tooth growth in guinea pigs.  Each animal received a particular dosage (0.5, 1, or 2 mg/day)
of vitamin C through one of the two specified delivery methods.  At the end of the period their
ondontoblast cell lengths were observed and recorded.

The data is formatted as follows:

```{r}
head(ToothGrowth)
```

#Analysis
##Data Exploration

We begin our analysis with a quick look at the data.  First, we look at the mean length of the 
cells by delivery method.

```{r}
aggregate(len ~ supp, data = ToothGrowth, FUN = mean)
```

It appears that delivering the vitamin C dosages via OJ may correspond with greater cell length independent
of dosage level.  

Lets also look at mean cell length by dose, regardless of delivery method.

```{r}
aggregate(len ~ dose, data = ToothGrowth, FUN = mean)
```

It also appears that increasing the dosage of vitamin C may lead to longer cells as well.

Finally, we look at mean cell length by both delivery method and dosage.

```{r}
aggregate(len ~ ., data = ToothGrowth, FUN = mean)
```

Somewhat confirming our suspicions above, it looks like OJ may lead to longer cells than VC and increasing
dosage levels of vitamin C may also lead to longer cells.  Of note is the observation that at a dosage
level of 2.0, it appears that it may not make a difference which delivery method was used.

Of course, our suspicions should be validated with some statistical analysis, which we will now perform.

##Statistical Analysis
For all of our statistical analysis we will make the assumption that our data is iid and normally
distributed.  Since we have a small sample size, 60, we will use Students T tests during our analysis.

First, we will consider the possibility that delivering the vitamin C via OJ yields longer cells than
delivering via VC.  We let our null hypothesis be that there is no significant difference, and let the
alternative hypothesis be that mean length is greater via OJ than with VC.

```{r, echo = FALSE}
length.vc = ToothGrowth[ToothGrowth$supp == 'VC',]$len
length.oj = ToothGrowth[ToothGrowth$supp == 'OJ',]$len
```
```{r}
result <- t.test(length.oj, length.vc, alternative = 'greater')
result$conf.int
result$p.value
```

We can see that the p-value for our above test is approximately 0.03 and is less than a significance level
of 0.05.

Next, we will consider the possibility that increasing dosages of vitamin C yield longer cells.  Here
we will have to make a few different comparisons.  In particular, we will look at difference is cell length
between dosages of 0.5 and 1.0, and 1.0 and 2.0.  In each case, the null hypothesis is that
there is no difference in mean cell length and the alternative is that the higher dose has a significantly
greater mean cell length. 

```{r, echo = FALSE}
length.0.5 = ToothGrowth[ToothGrowth$dose == 0.5,]$len
length.1.0 = ToothGrowth[ToothGrowth$dose == 1.0,]$len
length.2.0 = ToothGrowth[ToothGrowth$dose == 2.0,]$len
```

```{r}
# 0.5 vs 1.0
result <- t.test(length.1.0, length.0.5, alternative = 'greater')
result$conf.int
format(result$p.value, scientific = FALSE)
```
We can see that the p-value of this test is much smaller than a significance level of 0.05.


```{r}
# 1.0 vs 2.0
result <- t.test(length.2.0, length.1.0, alternative = 'greater')
result$conf.int
format(result$p.value, scientific = FALSE)
```
The same can be said of this p-value.

#Conclusions

In all of our tests, the p-values we obtained indicate that we can safely reject the null hypothesis
and accept the alternative.  This is of course conditioned on our assumptions of iid and normally distributed
data.  Thus, we can make the following conclusions.  

1. Delivering vitamin C to guinea pigs via orange juice leads to longer odontoblast cell length.
2. Increasing dosages from 0.5 to 1.0 and from 1.0 to 2.0 leads to longer odontoblast cell length.

A possible limitation of this analysis are that the beginning length of the cells was not available.
This could mean that each guinea pig began the study with different cell lengths which would pollute the
final cell length observation unless accounted for.
